---
title: "Exploratory Data Analysis: World Development Indicators"
author: "Emma Jinright"
date: "02-26-2025"
format: 
    revealjs:
        theme: dracula
        transition: fade
        scrollable: true
bibliography: references.bib
---

# Introduction

## World Development Indicators

This exploratory data analysis focuses on analyzing a few of the variables from the **World Development Indicators** dataset for 2022 made available by the World Bank [@worldbank].

## Variables

There are 14 variables:

:::{.columns}
:::{.column width="50%"}
 - `country`
 - `inflation_rate`
 - `exports_gdp_share`
 - `gdp_growth_rate`
 - `gdp_per_capita`
 - `adult_literacy_rate`
 - `total_population`
:::

:::{.column width="50%"}
 - `education_expenditure_gdp_share`
 - `measles_immunisation_rate`
 - `health_expenditure_gpd_share`
 - `income_inequality`
 - `unemployment_rate`
 - `life_expectancy`
 - `primary_school_enrolment_rate`
:::
:::

## Focus Variables

I will be focusing on 

:::{.incremental}
- **`total_population`**
- **`life_expectancy`**
- **`measles_immunisation_rate`**
:::

# Exploratory Data Analysis

# Total Population

## Average Population per Country

```{r}
#| label: EDA_1

pacman::p_load(tidyverse, readr, dplyr, ggplot2, skimr, scales, knitr, kableExtra)

wdi <- read_csv("./wdi.csv")

wdi_avg_pop <- wdi %>% 
  summarize(mean_total_population = mean(total_population),
            sd_total_population = sd(total_population))
kable(wdi_avg_pop, 
      col.names = c("Average Population", "Population Standard Deviation"), 
      format.args = list(big.mark = ","),
      align = 'c',)
```

##

```{r}
#| label: EDA_1.2

wdi_pop_rank <- wdi %>% 
  select(country, total_population) %>%
  arrange(desc(total_population))

kable(head(wdi_pop_rank, 5),
      caption = "Countries with the Largest Populations in 2022",
      col.names = c("Country", "Population"),
      format.args = list(big.mark = ","),
      align = 'c')
```

## 

```{r}
#| label: EDA_1.3

wdi_pop_rank_low <- wdi %>% 
  select(country, total_population) %>%
  arrange(total_population)

kable(head(wdi_pop_rank_low, 5),
      caption = "Countries with the Smallest Populations in 2022",
      col.names = c("Country", "Population"),
      format.args = list(big.mark = ","),
      align = 'c')

```

# Life Expectancy

## Average Life Expectancy per Country

```{r}
#| label: EDA_2

wdi_avg_life <- wdi %>% 
  
  summarize(mean_life_expectancy = mean(life_expectancy, na.rm = TRUE),
            sd_life_expectancy = sd(life_expectancy, na.rm = TRUE))
kable(wdi_avg_life, 
      col.names = c("Average Life Expectancy", "Life Expectancy Standard Deviation"), 
      align = 'c',
      digits = 2)
```

##

```{r}
#| label: EDA_2.2

wdi_life_rank <- wdi %>% 
  select(country, life_expectancy) %>%
  filter(!is.na(life_expectancy)) %>% 
  arrange(desc(life_expectancy))

kable(head(wdi_life_rank, 5),
      caption = "Countries with the Highest Life Expectancies in 2022",
      col.names = c("Country", "Life Expectancy"),
      align = 'c',
      digits = 2)
```

## 

```{r}
#| label: EDA_2.3

wdi_life_rank_low <- wdi %>% 
  select(country, life_expectancy) %>%
  filter(!is.na(life_expectancy)) %>% 
  arrange(life_expectancy)

kable(head(wdi_life_rank_low, 5),
      caption = "Countries with the Lowest Life Expectancies in 2022",
      col.names = c("Country", "Life Expectancy"),
      align = 'c',
      digits = 2)

```

# Measles Immunization Rate

This variable represents the **percentage of the population age 12 to 23 months that is vaccinated against measles** [@CDCmeasles].

## Average Immunization Rate per Country in Percentages

```{r}
#| label: EDA_3

wdi_avg_measles <- wdi %>% 
  summarize(mean_measles_immunisation_rate = mean(measles_immunisation_rate, na.rm = TRUE),
            sd_measles_immunisation_rate = sd(measles_immunisation_rate, na.rm = TRUE))
kable(wdi_avg_measles, 
      col.names = c("Average Immunization Rate", "Immunization Rate Standard Deviation"), 
      align = 'c',
      digits = 2)
```

##

```{r}
#| label: EDA_3.2

wdi_measles_rank <- wdi %>% 
  select(country, measles_immunisation_rate) %>%
  filter(!is.na(measles_immunisation_rate)) %>% 
  arrange(desc(measles_immunisation_rate))

kable(head(wdi_measles_rank, 5),
      caption = "Countries with the Highest Percentage of Measles Immunization in 2022",
      col.names = c("Country", "Percent Immunization for Measles"),
      align = 'c',
      digits = 2)
```

## 

```{r}
#| label: EDA_3.3

wdi_measles_rank_low <- wdi %>% 
  select(country, measles_immunisation_rate) %>%
  filter(!is.na(measles_immunisation_rate)) %>% 
  arrange(measles_immunisation_rate)

kable(head(wdi_measles_rank_low, 5),
      caption = "Countries with the Lowest Percentage of Measles Immunization in 2022",
      col.names = c("Country", "Percent Immunization for Measles"),
      align = 'c',
      digits = 2)

```

## Measles Protection

According to the United States' Center for Disease Control and Prevention (CDC), a vaccination rate of 95% of both doses of the measles vaccine means that the population is protected from outbreaks.

```{r}
#| label: EDA_3.4

wdi_measles <- wdi_measles_rank %>% 
  summarize(n_countries = n(),
            n_countries_95 = sum(measles_immunisation_rate >= 95),
            percent_95 = (sum(measles_immunisation_rate >= 95)/n())*100)

kable(wdi_measles,
     col.names = c("Number of Countries", "Number of Countries with Measles Immunization Rates At or Above 95%", "Percentage of Countries with Measles Immunization Rates At or Above 95%"),
     align = 'c',
     digits = 2)
```

# Figures & Tables

##

```{r}
#| label: fig-pop
#| echo: false
#| message: false
#| warning: false
#| fig.cap: "This is a bar graph displaying the total population for each country in the top 15% of populations by measles immunization rate (Source: @worldbank)."

wdi_analysis_1 <- wdi %>% 
   filter(!(is.na(measles_immunisation_rate)) & (total_population >= quantile(total_population, 0.85)) ) %>% 
  arrange(desc(total_population)) %>% 
  mutate(vac_95 = ifelse(measles_immunisation_rate >= 95, "Protected", "Unprotected"),
         country = factor(country, levels = country)) %>% 
  select(country, total_population, life_expectancy, measles_immunisation_rate, vac_95)

ggplot(data = wdi_analysis_1, aes(x=country,y=total_population, group=vac_95, fill=vac_95)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(
    breaks = seq(0, 2500000000, by = 100000000),
    labels = label_comma(scale = 1e-9, suffix = "B")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
  labs(title = "Total Population of Countries by Measles Vaccination Percentage", x = "Country", y = "Total Population (in billions)")

```

##

```{r}
#| label: fig-life
#| echo: false
#| message: false
#| warning: false
#| fig.cap: "This is a bar graph displaying the life expectancy for each country in the top 15% of populations by measles immunization rate (Source: @worldbank)."

wdi_analysis_2 <- wdi %>% 
   filter(!(is.na(measles_immunisation_rate)) & (total_population >= quantile(total_population, 0.85)) ) %>% 
  arrange(desc(life_expectancy)) %>% 
  mutate(vac_95 = ifelse(measles_immunisation_rate >= 95, "Protected", "Unprotected"),
         country = factor(country, levels = country)) %>% 
  select(country, total_population, life_expectancy, measles_immunisation_rate, vac_95)

ggplot(data = wdi_analysis_2, aes(x=country,y=life_expectancy, group=vac_95, fill=vac_95)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotating x-axis labels for better readability
  labs(title = "Life Expectancy of Countries by Measles Vaccination Percentage", x = "Country", y = "Life Expectancy")

```

## 

```{r}
#| label: fig-scatter
#| echo: false
#| message: false
#| warning: false
#| fig.cap: "This is a scatter plot showing the relationship between total population and life expectancy by measles immunization rate (Source: @worldbank)."

wdi_analysis_3 <- wdi %>% 
   filter(!(is.na(measles_immunisation_rate))) %>% 
  arrange(desc(total_population)) %>% 
  mutate(vac_95 = ifelse(measles_immunisation_rate >= 95, "Protected", "Unprotected"),
         country = factor(country, levels = country)) %>% 
  select(country, total_population, life_expectancy, measles_immunisation_rate, vac_95)

ggplot(data = filter(wdi_analysis_3, !is.na(life_expectancy)), aes(x=life_expectancy,y=total_population, group=vac_95, color=vac_95)) +
  geom_point() +
  scale_y_continuous(
    breaks = seq(0, 2500000000, by = 100000000),
    labels = label_comma(scale = 1e-9, suffix = "B")) +
  scale_x_continuous(
    breaks = seq(50, 85, by = 5),
    limits = c(50, 85)) +
  labs(title = "Total Population vs Life Expectancy by \nMeasles Vaccination Percentage", x = "Life Expectancy (in years)", y = "Total Population (in billions)")

```

##

```{r}
#| label: tbl-overview
#| echo: false
#| message: false
#| warning: false

overview_data <- wdi %>% 
     filter(!(is.na(measles_immunisation_rate)) & !(is.na(life_expectancy))) %>% 
  mutate(vac_95 = ifelse(measles_immunisation_rate >= 95, "Protected", "Unprotected")) %>% 
  group_by(vac_95) %>% 
  summarize(average_population = mean(total_population),
            sd_population = sd(total_population),
            average_life_expectancy = mean(life_expectancy),
            sd_life_expectancy = sd(life_expectancy))

kable(overview_data, caption = "Population and Life Expectancy of Countries by Protection Against Measles",
      digits = 2,
      col.names = c("Is the Population Protected Against Measles?", "Average Population", "Population Standard Deviation", "Average Life Expectancy", "Life Expectancy Standard Deviation"),
      format.args = list(big.mark = ","),
      align = 'c',
      booktabs = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#2f4f4f") %>% 
  column_spec(1, bold = TRUE)

```

# Conclusions

##
- Good overview of statistics surrounding total population, life expectancy, and measles immunization rates for countries in 2022
- Future Steps:
    - Regression Analysis to determine the significance of relationships described above

## References
